General linear Models Coursework

### 2022-11-18 Draft_final.Rmd
Add files via upload

## Part 2: Regressing expenditure on income
Fitting the model $y_{i}=\beta_{0}+\beta_{1}x_{i}+\epsilon$ where $y$ is the weekly household expenditure and the explanatory variable $x$ is weekly income and $i \in \{1,...,1200\}$ is the ith household in the dataset produces the below table (these are the same for all linear models mentioned in parts 2 to 6).It shows the estimated values of the intercept ($\beta_0$) and the estimated 'effect' of a 1 pound increase of income on expenditure ($\beta_1$) and their associated standard errors.
```{r,echo=F}
#Question 2
#Fitting linear model
lm.income<-lm(expenditure~income,data = expend.df)
#creating table to display results
knitr::kable(as.data.frame(tidy(lm.income)[,1:3]),digits=3)%>%
  kable_styling(position = "center",latex_options = "HOLD_position")
```

```{r,echo=F,fig.height=4.2}
#Creating diagnostic plots
par(mfrow=c(1,2))
plot(lm.income,1)
plot(lm.income,2)
```
\newpage

@@

#Producing table of results
knitr::kable(as.data.frame(tidy(lm.income.sqd)[,1:3]),digits=3)%>%
  kable_styling(position = "center",latex_options = "HOLD_position")
```

@@

#Producing table of results
knitr::kable(as.data.frame(tidy(lm.log.income)[,1:3]),digits=3)%>%
  kable_styling(position = "center",latex_options = "HOLD_position")
```

@@

#Producing table of results
knitr::kable(as.data.frame(tidy(lm.log.income.sqd)[,1:3]),digits=3)%>%
  kable_styling(position = "center",latex_options = "HOLD_position")
```

```{r,echo=F,fig.height=8}
#Creating diagnostic plots
par(mfrow=c(2,2))
plot(lm.log.income.sqd)
```

@@

\newpage
## Part 6: Comparing models
Comparing the models from part 2-5 using AIC and adjusted $R^2$:

@@

```{r,echo=F}
#Producing table of results
knitr::kable(as.data.frame(tidy(lm.log.income.sqd)[,1:3]),digits=3)%>%
  kable_styling(position = "center",latex_options = "HOLD_position")
```

@@

Above is a table of the exponentatied estimated coefficients of all the variables which were approximately 1. These are all the possible interaction levels (excluding the dummy level) of the income interactions. Thus all income interactions are not meaningful/important as their exponentiated value is approximately 1. Therefore, their effect on expenditure is negligible so I remove them from the model as a simpler model is preferred. After removing these 3 variables, removing any more does not further decrease the AIC thus this simplified model is my final model.
```{r,include=F}
#updating to simpler model
lm.simple.model.1 <- update(lm.full.1.deletion, .~.-income:lab.force-income:hh.size-income:house.ten)

@@

#Therefore this is my final model
lm.final.model <- lm.simple.model.1
```

```{r,echo=F,fig.height=6}
#Producing diagnostic plots
par(mfrow=c(2,2))
plot(lm.final.model)
```

Producing diagnostic plots for this final model shows no issues with the model fit. The residuals display a random scattering of points indicating the linearity assumption has been met and the residuals variance does not appear to increase/decrease except for a few values but I don't think this is enough evidence to say that the homoscedasticity assumption is invalid. This is further backed up by a straight nearly horizontal red line in the scale-location plot with an even random scattering of residuals around it. The points follow a straight line in the QQ plot with the tails deviating slightly but not extreme enough to indicate the normality assumption is invalid. The cooks distance plot shows no influential points are present in the data.

## Part 8: Inference of final model
The final model is $log(y_{ijkwz})=\beta_{0}+\beta_{1}x_{ijkwz}+\beta_{2}x_{ijkwz}^{2}+\alpha_{j}^{G}+\alpha_k^{S}+\alpha_{w}^{T}+\alpha_{z}^{L}+\alpha^{SG}_{jk}+\epsilon_{ijkwz}$.

>>>>

## Part 2: Regressing expenditure on income
Fitting the model $y_{i}=\beta_{0}+\beta_{1}x_{i}+\epsilon$ where $y$ is the weekly household expenditure and the explanatory variable $x$ is weekly income and $i \in \{1,...,1200\}$ is the ith household in the dataset produces the below table (these are the same for all linear models mentioned in parts 2 to 6).It shows the estimated values of the intercept ($\beta_0$) and the estimated 'effect' of a 1 pound increase of income on expenditure ($\beta_1$) and their associated standard errors.
```{r,echo=F}
#Question 2
#Fitting linear model
lm.income<-lm(expenditure~income,data = expend.df)
#creating table to display results
knitr::kable(as.data.frame(tidy(lm.income)[,1:3]),digits=3,caption = "Estimated coefficients and SE for part 2 model")%>%
  kable_styling(position = "center",latex_options = "HOLD_position")
```

```{r,echo=F,fig.height=3.8}
#Creating diagnostic plots
par(mfrow=c(1,2))
plot(lm.income,1)
plot(lm.income,2)
```
\newpage

@@

#Producing table of results
knitr::kable(as.data.frame(tidy(lm.income.sqd)[,1:3]),digits=3,caption = "Estimated coefficients and SE for part 3 model")%>%
  kable_styling(position = "center",latex_options = "HOLD_position")
```

@@

#Producing table of results
knitr::kable(as.data.frame(tidy(lm.log.income)[,1:3]),digits=3,caption = "Estimated coefficients and SE for part 4 model")%>%
  kable_styling(position = "center",latex_options = "HOLD_position")
```

@@

#Producing table of results
knitr::kable(as.data.frame(tidy(lm.log.income.sqd)[,1:3]),digits=3,caption = "Estimated coefficients and SE for part 5 model")%>%
  kable_styling(position = "center",latex_options = "HOLD_position")
```

```{r,echo=F,fig.height=6}
#Creating diagnostic plots
par(mfrow=c(2,2))
plot(lm.log.income.sqd)
```

@@

## Part 6: Comparing models
Comparing the models from part 2-5 using AIC and adjusted $R^2$:

@@

```{r,echo=F}
#Producing table of results
knitr::kable(as.data.frame(tidy(lm.log.income.sqd)[,1:3]),digits=3,caption = "Estimated coefficients and SE for part 5 model")%>%
  kable_styling(position = "center",latex_options = "HOLD_position")
```

@@

Above is a table of the exponentatied estimated coefficients of all the variables which were approximately 1. These are all the possible interaction levels (excluding the dummy level) of the income interactions. Thus all income interactions are not meaningful/important as their exponentiated value is approximately 1. Therefore, their effect on expenditure is negligible so I remove them from the model as a simpler model is preferred. After removing these 3 variables, removing any more does not further decrease the AIC, furthermore the difference in AIC is small with the simple models AIC being **1406.732** compared to the more complicated model with an AIC of **1386.878**. Thus this simplified model is my final model.
```{r,include=F}
#updating to simpler model
lm.simple.model.1 <- update(lm.full.1.deletion, .~.-income:lab.force-income:hh.size-income:house.ten)

@@

#Therefore this is my final model
lm.final.model <- lm.simple.model.1

AIC(lm.final.model)
```
Producing diagnostic plots for this final model (seen below) shows no issues with the model fit. The residuals display a random scattering of points indicating the linearity assumption has been met and the residuals variance does not appear to increase/decrease except for a few values but I don't think this is enough evidence to say that the homoscedasticity assumption is invalid. This is further backed up by a straight nearly horizontal red line in the scale-location plot with an even random scattering of residuals around it. The points follow a straight line in the QQ plot with the tails deviating slightly but not extreme enough to indicate the normality assumption is invalid. The cooks distance plot shows no influential points are present in the data.
```{r,echo=F,fig.height=6}
#Producing diagnostic plots
par(mfrow=c(2,2))
plot(lm.final.model)
```

## Part 8: Inference of final model
Let $y$ denote the expenditure of the household, $x$ the income of the household, $\alpha^{G}$ the sex of the household head, $\alpha^{S}$ the size of the household, $\alpha^{T}$ the tenure type, $\alpha^{L}$ the labour force and $\alpha^{SG}$ the interaction between sex of household head and size of household. Then the final model is:
\begin{equation*}
log(y_{ijkwz})=\beta_{0}+\beta_{1}x_{ijkwz}+\beta_{2}x_{ijkwz}^{2}+\alpha_{j}^{G}+\alpha_k^{S}+\alpha_{w}^{T}+\alpha_{z}^{L}+\alpha^{SG}_{jk}+\epsilon_{ijkwz}
\end{equation*}
Where $\alpha_{iM}^{G}=\alpha_{i1}^{S}=\alpha_{iPriv}^{T}=\alpha_{iFT}^{L}=\alpha^{SG}_{iMk}=\alpha^{SG}_{ij1}=0$ where $j \in \{M,F\}, k \in \{1,2,3,4,5+\}, w \in \{Pub,Priv,Own\}, z \in \{FT,PT,U\}$ and $i \in \{1,...,n_{jkwz}\}$ with $n_{jkwz}$ number households with the $j^{th}$ sex of household head, $k^{th}$ number of persons in the household, $w^{th}$ tenure type and $z^{th}$ labour force type.
\newline

Due to the model being a log-linear model the explanatory variables have a multiplicative effect on the response. Exponentiating the estimated coefficients to find their effect on expenditure gives the below table. I note that all values have been rounded to 3 decimal places thus the zero p-values are due to p-values being too small be displayed and similarly the effect of income squared is too small to be shown. My follwoing inference will be based on this table.

```{r,echo=F}
#Calculating confidence intervals

lwr <-exp(tidy(lm.final.model)$estimate+c(-1)*qnorm(0.975)*tidy(lm.final.model)$std.error)
upr <- exp(tidy(lm.final.model)$estimate+c(1)*qnorm(0.975)*tidy(lm.final.model)$std.error)

Point.estimates <- round(exp(tidy(lm.final.model)[,2]),3)
p.values <- round(tidy(lm.final.model)[,5],3)

#Producing tables to display result
knitr::kable(data.frame("coefficent"=tidy(lm.final.model)[,1],
                 Point.estimates,
                 "Confidence Interval 95%"=c(paste("(",round(lwr,3),", ",round(upr,3),")")),
                 p.values),digits=3,caption = "Summary of exponentiated variables affecting expeniture")%>%
  kable_classic(full_width=F,latex_options = "HOLD_position",position= "center")

```
First, interpreting intercept value, it is for a male headed household of 1 person, publicly rented in full time employment who has an income of £0. Clearly this does not make sense. The difference between full time work and being unemployed was highly statistically significant. With those unemployed having an associated decrease in expenditure of (*8.5%*,*34.1%*) 95% CI compared to those in full time work with all other variables held constant. However the difference between full time work and being inactive or in part time work was not significant (CI included 1) therefore with all other variables constant there is no evidence that being in one of these 3 factors compared to another influences the expenditure of the household.

The difference between publicly rented properties expenditure and either privately rented or owned properties expenditure were statistically significant. With the associated increase in expenditure of (*22%*,*44.1%*) and (*11.5%*,*28.1%*) 95% CI's respectively with all other variables held constant.

Due to squared income squared the effect of an increase in income on expenditure depends on the initial value of income. Assuming all other variables are held constant the percentage increase o expenditure is $(\frac{exp(\beta_1(x+y)+\beta_2(x+y)^2)}{exp(\beta_1x+\beta_2x^2)}-1)*100$ where x is the initial value of income and y is the increase in income, $\beta_1=0.002285$ and $\beta_2=-0.000001036$. Thus, for a household earning the mean income of £512, a £20 increase in income increases expenditure by approximately *2.4%* whereas for a household earning only £300 the same £20 increase, increases expenditure by approximately *3.3%*. Due to the negative quadratic term the higher the initial income the less of a positive multiplicative effect an increase in expenditure has on expenditure.

```{r,include=F}
#calculating percent increase
beta_1 <- 0.002285
beta_2 <- -0.000001036
x <- 300
inc <- 20

(exp(beta_1*(x+inc)+beta_2*(x+inc)^2)/exp(beta_1*x+beta_2*x^2)-1)*100
```

The final 2 variables, sex of household head and household size have an interaction between them. However, only the household size of 3 level has a statistically significant interaction with sex of household head. Therefore as the other levels

The sex of the household head is highly statistically significant with being female head increasing the expenditure by (*5.6%*,*25.9%*) 95% CI compared to a male head with all other variables held constant. Household size is extremely statistically significant with an increase in expenditure of (*22.4%*,*45.5*),  (*31.3%*,*66.8%*), (*34.9%*,*81.6%*) 95% CI's for households of size 2,4,5+ when compared to a household of size 1 with all other variables held constant. For the effect on expenditure between a household of size 1 and size 3 because of the significant interaction with sex if the household head is male the CI is (*36.2%*,*71.9%*) and (*-15.7%*,*+52.3%*) if female is it is with all other variables held constant. Note that the for the female head CI 0% is included therefore there is no statistically significant difference in expenditure between a household of size 1 with male head and a household of size 3 with a female head with all other variables held constant.

Overall an increase in income, higher household size, being a female head and living in a owned or privately rented increase the % expenditure when compared to a single male headed household living in a publicly rented house (unless a female head with size 3 household). Whereas being unemployed compared to in full time work decreases the % of expenditure.

All the variables mentioned above were statistically significant in their association with expenditure. However, just because they are significant, it does not mean they influence expenditure. There could be confounding or simple chance causing these results. However, some areas of confidence are:

* income, logically if you make more money you have more available to spend 
* household size, larger households are more expensive thus require more spending
